---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
  
```{r, include = FALSE}
require(knitr)
knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "500px",
  dpi = 150,
  fig.align = "center",
  fig.width=4,
  fig.height=4,
  out.width=400
)
```

# trees[within](http://github.com/PoonLab/twt)trees

**trees[within](http://github.com/PoonLab/twt)trees** (`twt`) is an R package for running [phylodynamic](https://en.wikipedia.org/wiki/Viral_phylodynamics) simulations.

Specifically, **twt** uses exact stochastic simulation to generate forward-time population dynamics under a [compartmental model](https://en.wikipedia.org/wiki/Compartmental_models_(epidemiology)), followed by the reverse-time simulation of a transmission history (outer tree) given these dynamics, and a pathogen phylogeny (inner tree) nested within that transmission history.


#### Features

* Use R expressions to specify rates; for example:
  * Constant expressions like `0.1` or `beta*S*I`
  * Probability distributions like `1+rpois(1,0.5)` for stochastic transmission bottlenecks
  * Time-varying rates like `beta0*(1+beta1*cos(w*t))` for periodic variation in transmission rates.
* Several built-in functions for visualizing your results.
* Accessible simulation outputs: All events are recorded in data frames for visualization and analysis.
* Your model can have repeated infections of the same host (superinfection), which induce discordant inner trees.
* Easy to install and use: **twt** is implemented exclusively in R.


### Installation

The recommended method to install this package is to download a release or clone the repository, navigate to that directory and then run the `INSTALL` command:
```bash
git clone https://github.com/PoonLab/twt.git
R CMD INSTALL twt
```
The last command can be run as `sudo` if you want the package to be available for all users.

Alternatively, you may use the [devtools](https://cran.r-project.org/web/packages/devtools/index.html) package to automate some of the above:
```{r eval=FALSE}
require(devtools)
devtools::install_github("PoonLab/twt")
```
However, `devtools` requires a large number of other packages to be installed.


### Usage

If you are in a hurry, you can use pipes to quickly generate a tree in one line:
```{r eval=FALSE, message=FALSE}
require(twt)
phy <- Model$new(read_yaml("examples/SIR_serial.yaml")) |> sim.dynamics() 
       |> sim.outer.tree() |> sim.inner.tree() |> as.phylo()
```
However, you may want to generate replicate simulations at different stages of this simulation, or visualize the outputs.


The following is a step-by-step **twt** workflow under an [SIR model](https://en.wikipedia.org/wiki/Compartmental_models_(epidemiology)#SIR_model) with serial sampling and three demes.

**twt** uses the [YAML](https://en.wikipedia.org/wiki/YAML) markup language to specify a model:
```{r message=FALSE}
require(twt, quietly = TRUE)
settings <- read_yaml("examples/migration.yaml")
```

```{r echo=FALSE}
# reduce overwrite of figures in repo
set.seed(3)
```

This model specification (which is now a `list` object in R) is used to create a Model object that can be visualized as a network, which is a convenient means of troubleshooting your model:
```{r out.width="60%", fig.width=5}
mod <- Model$new(settings)
plot(mod)
```


Next, we simulate the population dynamics from this model forward in time.  The results can be visualized with a call to the generic `plot` method:
```{r out.width="60%", fig.width=5, fig.height=4}
dynamics <- sim.dynamics(mod)
plot(dynamics, ylim=c(0, 200))
```

To simulate an "outer" transmission tree, we pass the `Dynamics` object to another function.
```{r}
outer <- sim.outer.tree(dynamics)
```

As before, **twt** provides a generic plot function for visualizing the simulation results.  Each horizontal line segment in this plot represents an infected host (grey if unsampled), and each vertical arrow represents a transmission event:
```{r fig.width=4, fig.height=5}
plot(outer)
```

If you choose to stop at this point, you can convert the OuterTree object to a phylogeny compatible with the R package `ape`, and exported with `write.tree`:
```{r echo=FALSE}
par(mar=c(1,1,1,1))
```

```{r out.width="40%"}
phy <- as.phylo(outer)
plot(phy)
```

All simulated events are embedded in this `phylo` object, making it easier to annotate the tree.  In this example, we are colouring the branches to indicate which compartment (deme) is occupied in the transmission history:
```{r out.width="50%"}
L <- tree.layout(phy)
pal <- c(I1='firebrick', I2='orange', I3='cadetblue')
plot(L, type='n', mar=c(3,1,1,5))
lines(L, col=pal[L$edge$compartment], lwd=2)
```
Note we can change colours mid-branch because we know exactly when the migrations occur.


To simulate an inner tree within the outer tree, we pass the `OuterTree` object to another function.  The resulting `InnerTree` object can also be converted to an `ape::phylo` object for viewing and writing:
```{r out.width="50%"}
inner <- sim.inner.tree(outer)
L <- tree.layout(as.phylo(inner))
L$nodes$label <- L$nodes$host  # use Host labels instead of Pathogen
plot(L, type='n', label='t')
lines(L, col=pal[L$edge$compartment], lwd=2)
```
Because this particular model (`migration.yaml`) has incomplete bottlenecks, we see some discordance between the inner and outer tree topologies.


#### Funding

Development of *treeswithintrees* was supported by the Government of Canada through [Genome Canada](https://www.genomecanada.ca/) and the [Ontario Genomics Institute](https://www.ontariogenomics.ca/) (OGI-131), and by the [Canadian Institutes of Health Research](http://cihr-irsc.gc.ca/e/193.html) (PJT-155990).