---
title: "An introduction to twt"
author: "Art Poon"
date: "4/21/2020"
output: rmarkdown::html_vignette
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*treeswithintrees* (twt) is an R package for simulating the proliferation of pathogen (parasite) lineages that are being transmitted between individual hosts, or between compartments within a single host.
Put another way, *twt* uses discrete-event simulation methods to generate random outcomes from a nested branching process defined by the user.
*twt* performs a simulation in two stages.
First, 

## Installation

If you have already installed the `devtools` package in your R framework, then the easiest way to install *twt* is to use the following commands:
```{r}
require(devtools)
devtools::install_github('PoonLab/twt')
```

However, the `devtools` package installs a large number of package dependencies, some of which have to be compiled locally.  Some users may prefer to 


## Basic example

First, we load the package:
```{r}
require(twt)
```
