---
output: 
  revealjs::revealjs_presentation:
    css: style.css
    center: true
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## *treeswithintrees*: an R package for exact discrete-event simulation of virus phylogenies

**Art Poon**  
Western University, Canada


## Phylodynamics and simulation

* [Phylodynamics](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1002947) is the reconstruction of epidemiological and immunological processes from the shape of the virus phylogeny.

* Simulation is an important technique to:
  * quickly characterize a model
  * validate methods to fit a model to data
  * fit a model with likelihood-free methods


## State of the art

* Several programs that enable the user to simulate trees under customized dynamic models.
  * [phydynR](https://github.com/emvolz-phylodynamics/phydynR) uses coalescent simulation given deterministic solution of dynamic system.
  * [phylodyn](https://github.com/mdkarcher/phylodyn) simulates coalescent trees under arbitrary effective population size trends.
  * [MASTER](http://tgvaughan.github.io/MASTER/) provides a flexible language for the forward discrete-event simulation of models.
  * [TreeModelAdequacy](https://github.com/sebastianduchene/tree_model_adequacy) uses simulation to assess whether a phylodynamic model can reproduce features of the data.


## Unmet needs

* Issues with approximate numerical solution of phylodynamic models as systems of differential equations. 
* Exploring stochastic dynamics of systems (deterministic solutions too limited).
* More flexible model specification: *e.g.*, arbitrary coalescent rate variation within hosts; variation in host attributes over time.
* A more feature-rich visualization toolkit.


## Proposed workflow

* Forward-time simulation of compartmental dynamics:
  * transmission - movement of virus to uninfected host
  * migration - movement of virus to infected host (superinfection)
  * transition - a given host changes between states
  
* Reverse-time simulation of (outer) transmission tree relating sampled hosts
* Coalescent simulation of (inner) virus phylogeny 

## Object classes

* `Model`
  * `CompartmentType` - *e.g.*, susceptible hosts
  * `Compartment` - individual instances of a given `CompartmentType`
  * `Lineage` - a sampled lineage of the virus tree
* `Run` - derived class of `Model` that contains outer tree
* `EventLogger` - data frame of both outer and inner tree events


## Model specification with YAML

```yaml
InitialConditions:
  originTime: 10.0
  size:
    host: 100
  indexType: 'host'
CompartmentTypes:
  'host':
    branching.rates: (host=0.01)
    effective.size: 100.0
    generation.time: 0.01
    bottleneck.size: 1
Compartments:
  'I':
    type: host
    replicates: 10
Lineages:
  'I':
    type: 'virus'
    sampling.time: 0             
    location: I
    replicates: 3     
```


## Example

* SI model with 2 host types, 5 lineages sampled from each of 10 hosts

```{r echo=TRUE, message=FALSE}
require(twt)
path <- system.file('extdata', 'structSI.yaml', package='twt')
mod <- Model$new(yaml.load_file(path)); mod
```


## Outer tree
```{r echo=TRUE, fig.width=10, fig.height=5}
set.seed(1); run <- sim.outer.tree(mod)
par(mfrow=c(1,2)); plot(run, 's'); plot(run)
```

## Piece-wise linear coalescent model
```{r}


```


## Inner tree
```{r echo=TRUE, fig.height=5, fig.width=5}
phy <- sim.inner.tree(run)
plot(phy, cex=0.8)
```
